Episodic memory: mental time travel or a quantum `memory wave' function?

### 2021-01-12 paper/main.tex
updated bibliography, draft of edits to address R1 comments, updated diff files

One practical way to distinguish content from context representations is to separate out mental (or neural) representations that drift rapidly \citep[content;][]{PolyEtal05a, MannEtal12} or gradually \citep[context;][]{PolyEtal05a, MannEtal11, HowaEtal12, LohnEtal18, LongKaha18, FolkEtal18}.  An intuition for why contextual representations might be expected to evolve gradually is laid out by \cite{PolyKaha08}.  Essentially, slowly drifting thoughts may be used to ``index'' more rapidly drifting thoughts that are temporally proximal.  This notion, that our brains maintain parallel mental representations drifting at different timescales, has been well-characterized in a series of elegant fMRI and ECoG studies by Uri Hasson and colleagues~\citep{HassEtal08, LernEtal11, HoneEtal12a, AlyEtal18}.  Subsequent work by the same group has shown that this hierarchy of drifting mental representations plays a central role in how we remember continuous experiences and segment our experiences into discrete events~\citep{BaldEtal17}.  This body of work suggests that different brain regions reflect \textit{temporal receptive windows} that characterize the timescale(s) to which each region is maximally sensitive.  This research complements work on the \textit{temporal window of integration} in the speech and perception literatures~\citep[e.g.,][]{ViemWake91, GautEtal12, vanWEtal07}.  The temporal window of integration refers to the duration over which physically distinct stimuli are treated (by some brain system or network) as a single percept or atomic unit.  This systems-level work mirrors the discovery of \textit{time cells} in the rodent hippocampus that respond preferentially to a specific time relative to a temporal reference (e.g., time elapsed since starting to run on an exercise wheel), whereby each cell appears to integrate incoming information at a different rate~\citep{PastEtal08, MacDEtal11} and the population activity serves to represent information at a range of timescales~\citep{MauEtal18}. Other work suggests that the lateral entorhinal cortex also plays a role in integrating information across a wide range of timescales~\citep{TsaoEtal18}, supporting precise temporal recall~\citep{MontEtal19}.  In addition to this work suggesting that individual systems or neurons maintain representations at their preferred timescales, a number of studies also report neurons and small populations exhibiting \textit{temporal multiplexing}, whereby information at different timescales is reflected in different features of an individual neuron's (or population's) activity patterns, such as spike timing versus firing rate or local field potential oscillations at different frequencies~\citep{CaruEtal18, KnigEich13, WatrEtal13, PanzEtal10}.

The above ideas about parallel neural representations that reflect information at multiple timescales have been formalized in a family of theoretical models over the past several decades~\citep{HowaKaha02a, DianEtal07, SedeEtal08, PolyEtal09, ShanEtal09, ShanHowa10, ShanHowa12, HowaEtal14, Rang18}.  In turn, these models were inspired in part by earlier theoretical work suggesting that temporally proximal experiences become linked through their shared contextual features~\citep{Este55a,AtkiShif68}.  Other recent work indicates that different aspects of the reinstatement process itself might also occur at multiple timescales.  For example, \cite{LindEtal19} found that high-level semantic information was reinstated prior to low-level (e.g., perceptual) details in an associative memory task.  Collectively, these experimental and theoretical studies make two important contributions to our understanding of how we remember our past.  First, the studies suggest that our ongoing experiences are ``blurred out'' in time by neural integrators (where different brain structures or sub-structures integrate with different time constants).  Second, the studies suggest that representations that reflect different timescales become bound together such that when we retrieve memories about our past, we reactivate representations that carry information at a range of timescales.  These reactivations explain how our brains reactivate prior contexts when we remember the past.

@@

One insight into this distinction comes from a recent study by \cite{HeusEtal18c}, which analyzed data collected by \cite{ChenEtal17} who had participants watch an episode of the BBC television show \textit{Sherlock} and then verbally recount what happened in the episode.  The authors used topic models~\citep{BleiEtal03} applied to human-generated annotations of each camera shot in the episode to characterize the episode's content with a timeseries of high-dimensional semantic feature vectors (see \textit{Defining the geometries of ongoing thoughts and memories}).  Each shot's annotation contained the following information: a brief narrative description of what was happening, the location where the scene took place, whether that location was indoors or outdoors, the names of all on-screen characters, the names of the character(s) who were in focus in the camera shot, the name(s) of any character(s) who spoke during the scene, the camera angle of the shot, a transcription of any on-screen text, and whether or not there was any music present in the auditory background~\citep[full details may be found in][]{ChenEtal17}.  The resulting \textit{topic vectors} describe the mix of ``themes'' present in each moment of the episode (Fig.~\ref{fig:embedding-models}B).  When characterized in this way, the episode's content exhibits periods of relative stability, followed by moments of rapid change. This interplay between stability and change is visible as ``blocks'' in the temporal correlation matrix (Fig.~\ref{fig:corrmats}A); \cite{HeusEtal18c} performed detailed analyses of these putative \textit{events} and how they are remembered.  Notably, the off-diagonal entries of the temporal correlation matrix (which reflect content similarity between temporally distant events in the episode) are nearly all close to 0.  This is surprising, because the episode itself comprises a murder mystery with recurring motifs that must be associated in the viewer's mind in order to make sense of the episode.  Nonetheless, when considering solely the semantic content of each moment of the episode, those long timescale associations are nearly entirely absent.  To the extent that this finding extends to real-world experiences, it comports with Heraclitus' notion that one cannot step in the same river twice~\citep{Hera}.

\begin{figure}[tp] \centering \includegraphics[width=0.8\textwidth]{figs/pres_rec_corrmats_sherlock} \caption{\textbf{Temporal correlation matrices of the dynamic content of a television episode and people's recalls of that episode.}  The panels of this figure are adapted from \cite{HeusEtal18c}, where full details may be found. \textbf{A. Temporal correlations exhibited by the content of the television episode.}  Each moment of an episode of the BBC television show \textit{Sherlock} has been characterized using a topic model~\citep{BleiEtal03} fit to detailed human-generated annotations between each scene cut.  The correlations between the topic vectors of a given pair of timepoints reflects the similarity in thematic content at those timepoints.  \textbf{B. Temporal correlations exhibited by the (average) content of participants' recalls of the episode.} The topic model used to characterize the episode in Panel A has been applied to each sentence from each participants' recall transcript.  The temporal correlation matrices describing each participants' recalls were re-sampled using cubic interpolation to have 100 rows and columns prior to averaging the resampled matrices across participants.  \textbf{C. Temporal correlations in the content of individual participants' recalls of the episode.} Each sub-panel displays the temporal correlation matrix for one experimental participant.}
\label{fig:corrmats}
\end{figure}

@@

\section*{The shapes of remembered experiences}
The study by \cite{HeusEtal18c} described above suggests one way of characterizing the dynamic content of complex naturalistic stimuli like television episodes.  The sequence of topic vectors over the course of the episode traces out a \textit{topic trajectory} through a high-dimensional feature space.  The sequence of topic vectors for a given participant's recalls traces out an analogous trajectory.  This characterization provides a geometric framework for comparing the \textit{shapes} that reflect how naturalistic experiences unfold over time, to the shapes of how we remember those experiences later (Fig.~\ref{fig:trajectories}).  One can then use the geometric transformations needed to map an experience's trajectory onto the trajectory of its later remembering to characterize the quality and content of memory.

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/trajectory}
\caption{\textbf{The shape of an unfolding experience.}  \textbf{A. The trajectory through topic space taken by a television episode.} The topic trajectory taken by a 50 minute television episode, projected onto two dimensions.  Each dot indicates an event identified using a Hidden Markov Model; the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B. The trajectory through topic space taken by participants' recalls of the television episode.} The average two-dimensional trajectory captured by 17 participants' recalls of the episode, with the same format and coloring as the trajectory in Panel A. The arrows reflect the directions of the average transitions taken by participants whose trajectories crossed that part of topic space; blue arrows denote directions of reliable agreement across participants.  Figure adapted from \cite{HeusEtal18c}, where additional details may be found.}
\label{fig:trajectories}
\end{figure}

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal18c} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal18c} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal18c} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experiences over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

@@

For real-world memory phenomena, and in many naturalistic memory paradigms, solving the matching problem is more complex.  People often experience real-world events (and scenes from movies or stories) as rich sensory experiences that derive deep meaning from surrounding moments, other prior experiences, emotions, goals, high-order understanding of the current situation, and so on.  Many of these aspects of our real-world experiences are not immediately verbalizable-- and even when we are able to describe some specific aspects of what we are experiencing, there are typically many possible ways we could have done so (e.g., varying in level of detail, specific wording, objectivity versus subjectivity, contextual details, etc.).  Further, because the way we experience each moment depends in part on our unique prior experiences, even ostensibly shared experiences can \textit{seem} different to each individual.  The way we recount a given experience is often a reflection of our own idiosyncratic experience of the original event, and is colored and distorted in an idiosyncratic way by our memory systems.  This distorted memory must then be verbalized or communicated, which is itself an idiosyncratic (and potentially lossy, in the information theoretic sense) process.  Solving the matching problem under these constraints poses a substantial theoretical and methodological challenge.  Specifically, matching requires building explicit models of the dynamic conceptual content of our experiences and memories, and then estimating when and how that experienced and remembered content aligns.  The approach featured in Figure~\ref{fig:trajectories} uses a word embedding space that is defined by a topic model~\citep{BleiEtal03}.  Each coordinate in this space reflects the conceptual content underlying a set of words, such that two different text passages that describe the same conceptual content (but using different words, levels of detail, etc.) are mapped onto nearby points in the word embedding space.  When these models are sufficiently expressive and robust to solve the matching problem, by modeling our experiences and memories within the same word embedding space, we can see a clear geometric alignment between the trajectories that our experiences (Fig.~\ref{fig:trajectories}A) and memories (Fig.~\ref{fig:trajectories}B) take through word embedding space~\citep[also see][]{HeusEtal18c}.

@@

Cutting-edge approaches to word embeddings that have achieved the best-to-date performance on modern natural language processing benchmarks are built using enormous deep neural networks often with many billions of parameters.  Because these models are very computationally expensive to fit, an important discovery has been that many layers in these models may be pre-trained without substantially affecting performance, even on new tasks.  One of the first of this new class of model, \textit{Bidirectional Encoder Representations from Transformers}~\citep[BERT;][]{DevlEtal18}, uses a pre-trained deep neural network, whose weights in a single output layer may be fine-tuned to suit a wide range of new tasks.  The latest additions to this class of models are called \textit{Generative Pre-trained Transformers}~\citep[GPT;][]{RadfEtal18, RadfEtal19, BrowEtal20}.

@@

One example of how the full distribution over possible mental states can provide insights into cognition may be seen in how the brain encodes spatial locations.  Theoretical~\citep[e.g., ][]{GersAbbo97} and empirical~\citep[e.g., ][]{PfeiFost13} work has demonstrated how place cell receptive fields may be modulated according to environmental features such as obstacles and the locations of rewarded goals.  In other words, although place cells encode an animal's location within an environment, even for a fixed cell and environment there is some variability with respect to where precisely within that environment the cell modulates its firing rate.  Further, a number of studies have suggested that some neurons respond to distinct non-contiguous places~\citep{FentEtal08, RichEtal14b, LeeEtal19, DerdEtal09, GrieEtal20} and times~\citep{PastEtal08}.  Taken together, these studies suggest that variability in the set of locations and/or times to which a neuron responds is not merely a reflection of noise or imprecision.  Rather, this variability might have some functional relevance in terms of facilitating the learning of paths around obstacles and towards goals.

@@

\section*{Acknowledgements}
I thank Luke Chang, Janice Chen, Paxton Fitzpatrick, Andrew Heusser, Christopher Honey, Caroline Lee, Talia Manning, Lucy Owen, Matthijs van der Meer, and Kirsten Ziman for feedback and scientific discussions. I also thank Janice Chen, Yuan Chang Leong, Christopher Honey, Chung Yong, Kenneth Norman, and Uri Hasson for sharing the data re-analyzed here~\citep{ChenEtal17}, and Paxton Fitzpatrick and Andrew Heusser for contributing analysis code that I adapted from a prior publication~\citep{HeusEtal18c}.  This work was supported in part by NSF EPSCoR Award Number 1632738. The content is solely my responsibility and does not necessarily represent the official views of my supporting organizations.

@@

\bibliographystyle{apa}
\bibliography{CDL-bibliography/memlab}
\end{document}

>>>>

One practical way to distinguish content from context representations is to separate out mental (or neural) representations that drift rapidly \citep[content;][]{PolyEtal05a, MannEtal12} or gradually \citep[context;][]{PolyEtal05a, MannEtal11, HowaEtal12, LohnEtal18, FolkEtal18}.  An intuition for why contextual representations might be expected to evolve gradually is laid out by \cite{PolyKaha08}.  Essentially, slowly drifting thoughts may be used to ``index'' more rapidly drifting thoughts that are temporally proximal.  This notion, that our brains maintain parallel mental representations drifting at different timescales, has been well-characterized in a series of elegant fMRI and ECoG studies by Uri Hasson and colleagues~\citep{HassEtal08, LernEtal11, HoneEtal12a, AlyEtal18}.  Subsequent work by the same group has shown that this hierarchy of drifting mental representations plays a central role in how we remember continuous experiences and segment our experiences into discrete events~\citep{BaldEtal17}.  This body of work suggests that different brain regions reflect \textit{temporal receptive windows} that characterize the timescale(s) to which each region is maximally sensitive.  This research complements work on the \textit{temporal window of integration} in the speech and perception literatures~\citep[e.g.,][]{ViemWake91, GautEtal12, vanWEtal07}.  The temporal window of integration refers to the duration over which physically distinct stimuli are treated (by some brain system or network) as a single percept or atomic unit.  This systems-level work mirrors the discovery of \textit{time cells} in the rodent hippocampus that respond preferentially to a specific time relative to a temporal reference (e.g., time elapsed since starting to run on an exercise wheel), whereby each cell appears to integrate incoming information at a different rate~\citep{PastEtal08b, MacDEtal11} and the population activity serves to represent information at a range of timescales~\citep{MauEtal18}. Other work suggests that the lateral entorhinal cortex also plays a role in integrating information across a wide range of timescales~\citep{TsaoEtal18}, supporting precise temporal recall~\citep{MontEtal19}.  In addition to this work suggesting that individual systems or neurons maintain representations at their preferred timescales, a number of studies also report neurons and small populations exhibiting \textit{temporal multiplexing}, whereby information at different timescales is reflected in different features of an individual neuron's (or population's) activity patterns, such as spike timing versus firing rate or local field potential oscillations at different frequencies~\citep{CaruEtal18, KnigEich13, WatrEtal13b, PanzEtal10}.

The above ideas about parallel neural representations that reflect information at multiple timescales have been formalized in a family of theoretical models over the past several decades~\citep{HowaKaha02a, DianEtal07, SedeEtal08, PolyEtal09, ShanEtal09, ShanHowa10, ShanHowa12, HowaEtal14, Rang19}.  In turn, these models were inspired in part by earlier theoretical work suggesting that temporally proximal experiences become linked through their shared contextual features~\citep{Este55a,AtkiShif68}.  Other recent work indicates that different aspects of the reinstatement process itself might also occur at multiple timescales.  For example, \cite{LindEtal19} found that high-level semantic information was reinstated prior to low-level (e.g., perceptual) details in an associative memory task.  Collectively, these experimental and theoretical studies make two important contributions to our understanding of how we remember our past.  First, the studies suggest that our ongoing experiences are ``blurred out'' in time by neural integrators (where different brain structures or sub-structures integrate with different time constants).  Second, the studies suggest that representations that reflect different timescales become bound together such that when we retrieve memories about our past, we reactivate representations that carry information at a range of timescales.  These reactivations explain how our brains reactivate prior contexts when we remember the past.

@@

One insight into this distinction comes from a recent study by \cite{HeusEtal21}, which analyzed data collected by \cite{ChenEtal17} who had participants watch an episode of the BBC television show \textit{Sherlock} and then verbally recount what happened in the episode.  The authors used topic models~\citep{BleiEtal03} applied to human-generated annotations of each camera shot in the episode to characterize the episode's content with a timeseries of high-dimensional semantic feature vectors (see \textit{Defining the geometries of ongoing thoughts and memories}).  Each shot's annotation contained the following information: a brief narrative description of what was happening, the location where the scene took place, whether that location was indoors or outdoors, the names of all on-screen characters, the names of the character(s) who were in focus in the camera shot, the name(s) of any character(s) who spoke during the scene, the camera angle of the shot, a transcription of any on-screen text, and whether or not there was any music present in the auditory background~\citep[full details may be found in][]{ChenEtal17}.  The resulting \textit{topic vectors} describe the mix of ``themes'' present in each moment of the episode (Fig.~\ref{fig:embedding-models}B).  When characterized in this way, the episode's content exhibits periods of relative stability, followed by moments of rapid change. This interplay between stability and change is visible as ``blocks'' in the temporal correlation matrix (Fig.~\ref{fig:corrmats}A); \cite{HeusEtal21} performed detailed analyses of these putative \textit{events} and how they are remembered.  Notably, the off-diagonal entries of the temporal correlation matrix (which reflect content similarity between temporally distant events in the episode) are nearly all close to 0.  This is surprising, because the episode itself comprises a murder mystery with recurring motifs that must be associated in the viewer's mind in order to make sense of the episode.  Nonetheless, when considering solely the semantic content of each moment of the episode, those long timescale associations are nearly entirely absent.  To the extent that this finding extends to real-world experiences, it comports with Heraclitus' notion that one cannot step in the same river twice~\citep{HeraCE}.

\begin{figure}[tp] \centering \includegraphics[width=0.8\textwidth]{figs/pres_rec_corrmats_sherlock} \caption{\textbf{Temporal correlation matrices of the dynamic content of a television episode and people's recalls of that episode.}  The panels of this figure are adapted from \cite{HeusEtal21}, where full details may be found. \textbf{A. Temporal correlations exhibited by the content of the television episode.}  Each moment of an episode of the BBC television show \textit{Sherlock} has been characterized using a topic model~\citep{BleiEtal03} fit to detailed human-generated annotations between each scene cut.  The correlations between the topic vectors of a given pair of timepoints reflects the similarity in thematic content at those timepoints.  \textbf{B. Temporal correlations exhibited by the (average) content of participants' recalls of the episode.} The topic model used to characterize the episode in Panel A has been applied to each sentence from each participants' recall transcript.  The temporal correlation matrices describing each participants' recalls were re-sampled using cubic interpolation to have 100 rows and columns prior to averaging the resampled matrices across participants.  \textbf{C. Temporal correlations in the content of individual participants' recalls of the episode.} Each sub-panel displays the temporal correlation matrix for one experimental participant.}
\label{fig:corrmats}
\end{figure}

@@

\section*{The shapes of remembered experiences}
The study by \cite{HeusEtal21} described above suggests one way of characterizing the dynamic content of complex naturalistic stimuli like television episodes.  The sequence of topic vectors over the course of the episode traces out a \textit{topic trajectory} through a high-dimensional feature space.  The sequence of topic vectors for a given participant's recalls traces out an analogous trajectory.  This characterization provides a geometric framework for comparing the \textit{shapes} that reflect how naturalistic experiences unfold over time, to the shapes of how we remember those experiences later (Fig.~\ref{fig:trajectories}).  One can then use the geometric transformations needed to map an experience's trajectory onto the trajectory of its later remembering to characterize the quality and content of memory.

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/trajectory}
\caption{\textbf{The shape of an unfolding experience.}  \textbf{A. The trajectory through topic space taken by a television episode.} The topic trajectory taken by a 50 minute television episode, projected onto two dimensions.  Each dot indicates an event identified using a Hidden Markov Model; the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B. The trajectory through topic space taken by participants' recalls of the television episode.} The average two-dimensional trajectory captured by 17 participants' recalls of the episode, with the same format and coloring as the trajectory in Panel A. The arrows reflect the directions of the average transitions taken by participants whose trajectories crossed that part of topic space; blue arrows denote directions of reliable agreement across participants.  Figure adapted from \cite{HeusEtal21}, where additional details may be found.}
\label{fig:trajectories}
\end{figure}

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal21} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal21} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal21} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people~\citep[also see][]{MortEtal20}, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experiences over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

@@

For real-world memory phenomena, and in many naturalistic memory paradigms, solving the matching problem is more complex.  People often experience real-world events (and scenes from movies or stories) as rich sensory experiences that derive deep meaning from surrounding moments, other prior experiences, emotions, goals, high-order understanding of the current situation, and so on.  Many of these aspects of our real-world experiences are not immediately verbalizable-- and even when we are able to describe some specific aspects of what we are experiencing, there are typically many possible ways we could have done so (e.g., varying in level of detail, specific wording, objectivity versus subjectivity, contextual details, etc.).  Further, because the way we experience each moment depends in part on our unique prior experiences, even ostensibly shared experiences can \textit{seem} different to each individual.  The way we recount a given experience is often a reflection of our own idiosyncratic experience of the original event, and is colored and distorted in an idiosyncratic way by our memory systems.  This distorted memory must then be verbalized or communicated, which is itself an idiosyncratic (and potentially lossy, in the information theoretic sense) process.  Solving the matching problem under these constraints poses a substantial theoretical and methodological challenge.  Specifically, matching requires building explicit models of the dynamic conceptual content of our experiences and memories, and then estimating when and how that experienced and remembered content aligns.  The approach featured in Figure~\ref{fig:trajectories} uses a word embedding space that is defined by a topic model~\citep{BleiEtal03}.  Each coordinate in this space reflects the conceptual content underlying a set of words, such that two different text passages that describe the same conceptual content (but using different words, levels of detail, etc.) are mapped onto nearby points in the word embedding space.  When these models are sufficiently expressive and robust to solve the matching problem, by modeling our experiences and memories within the same word embedding space, we can see a clear geometric alignment between the trajectories that our experiences (Fig.~\ref{fig:trajectories}A) and memories (Fig.~\ref{fig:trajectories}B) take through word embedding space~\citep[also see][]{HeusEtal21}.

@@

Cutting-edge approaches to word embeddings that have achieved the best-to-date performance on modern natural language processing benchmarks are built using enormous deep neural networks often with many billions of parameters.  Because these models are very computationally expensive to fit, an important discovery has been that many layers in these models may be pre-trained without substantially affecting performance, even on new tasks.  One of the first of this new class of model, \textit{Bidirectional Encoder Representations from Transformers}~\citep[BERT;][]{DevlEtal18}, uses a pre-trained deep neural network, whose weights in a single output layer may be fine-tuned to suit a wide range of new tasks.  The latest additions to this class of models are called \textit{Generative Pre-trained Transformers}~\citep[GPT;][]{RadfEtal19, BrowEtal20}.

@@

One example of how the full distribution over possible mental states can provide insights into cognition may be seen in how the brain encodes spatial locations.  Theoretical~\citep[e.g., ][]{GersAbbo97} and empirical~\citep[e.g., ][]{PfeiFost13} work has demonstrated how place cell receptive fields may be modulated according to environmental features such as obstacles and the locations of rewarded goals.  In other words, although place cells encode an animal's location within an environment, even for a fixed cell and environment there is some variability with respect to where precisely within that environment the cell modulates its firing rate.  Further, a number of studies have suggested that some neurons respond to distinct non-contiguous places~\citep{FentEtal08, RichEtal14b, LeeEtal19, DerdEtal09, GrieEtal20} and times~\citep{PastEtal08b}.  Taken together, these studies suggest that variability in the set of locations and/or times to which a neuron responds is not merely a reflection of noise or imprecision.  Rather, this variability might have some functional relevance in terms of facilitating the learning of paths around obstacles and towards goals.

@@

\section*{Acknowledgements}
I thank Luke Chang, Janice Chen, Paxton Fitzpatrick, Andrew Heusser, Christopher Honey, Caroline Lee, Talia Manning, Lucy Owen, Matthijs van der Meer, and Kirsten Ziman for feedback and scientific discussions. I also thank Janice Chen, Yuan Chang Leong, Christopher Honey, Chung Yong, Kenneth Norman, and Uri Hasson for sharing the data re-analyzed here~\citep{ChenEtal17}, and Paxton Fitzpatrick and Andrew Heusser for contributing analysis code that I adapted from a prior publication~\citep{HeusEtal21}, from which Figures~\ref{fig:corrmats} and~\ref{fig:trajectories} are also adapted.  I also note that I presented some of the work in this manuscript in a talk at the 2019 Context and Episodic Symposium.  This work was supported in part by NSF EPSCoR Award Number 1632738. The content is solely my responsibility and does not necessarily represent the official views of my supporting organizations.

@@

\bibliographystyle{apa}
\bibliography{CDL-bibliography/cdl}
\end{document}
